---
title: "Power analysis for multi-sample spatial transcriptomics with spaCraft"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Power analysis for multi-sample spatial transcriptomics with spaCraft}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning   = FALSE,
  message   = FALSE,
  # This vignette runs the full chain on the small bundled `mini_*` data:
  # every object is built in sequence starting from `visium_human_brain_pilot_data_list`,
  # all the way through one simulate -> recover -> test pass. Only the
  # Monte-Carlo tuning and power-grid sweeps are left unevaluated (eval = FALSE)
  # for build speed; flip them on (and raise n_sim) to run them.
  eval      = TRUE
)
```

## Overview

`spaCraft` provides a **pilot-based framework for statistical power analysis
and sample-size determination** in comparative, multi-sample spatial
transcriptomics studies. Given one or more annotated pilot samples, it learns a
cohort-level generative model and then estimates, by Monte Carlo, how many
samples per group are needed to detect an effect of a given size --- for two
distinct scientific questions:

- **SaLFC** --- a *spatially adjusted* differential-expression test for a
  Target-vs-Reference log-fold-change between two groups of samples; and
- **LOR** --- a baseline-anchored *compositional* test for a shift in the
  abundance of a target domain between groups.

The engine is a **generate--recover--test** pipeline. A three-layer generative
model is fitted from pilot data --- a baseline-anchored Binomial--logit model
for domain *composition*, a Fisher--Gaussian kernel mixture model (FGKMM) for
spatial *geometry*, and a Gaussian-process (NNGP) model for *expression*.
Synthetic cohorts are then simulated under a user-specified effect size, spatial
domains are recovered with a pilot-guided extension of BANKSY (`pBANSKY`), and
the chosen endpoint is applied. Repeating this over a grid of sample sizes
\eqn{K} and effect sizes yields empirical power, which is smoothed with a
monotone shape-constrained additive model to recommend a per-group \eqn{K}.

In this vignette we walk through the **entire workflow on human dorsolateral
prefrontal cortex (DLPFC) Visium data** (Maynard et al., 2021, obtained via
`spatialLIBD`), from a pilot list all the way to SaLFC and LOR power curves.

**Running this vignette.** Rather than loading a separate precomputed object at
each stage, this vignette **starts from the bundled, downsampled pilot list
`visium_human_brain_pilot_data_list` and builds every subsequent object in sequence** ---
`createspaCraftObject()` produces the working object, feature selection adds
the gene sets, the three estimators fill the parameter slots, the simulator
generates a cohort, and `pBANSKY()` recovers it. Each step simply consumes the
object returned by the previous one, so there is **no need to call**
`data(mini_obj_features)`, `data(mini_obj_fitted)`, and the like; those bundled
objects are just snapshots of this same chain at different points. The whole
pipeline through a single simulate \eqn{\to} recover \eqn{\to} test pass runs
live here, single-threaded (`n_cores = 1`). Two caveats follow from using a tiny
dataset: (i) because it is heavily downsampled (a few hundred genes and spots),
the feature-selection cutoffs are relaxed accordingly and the selected genes and
reported numbers are **illustrative, not the full-data results from the paper**;
and (ii) the Monte-Carlo **hyperparameter-tuning and power-grid sweeps** are
shown with small settings but left **unevaluated** (`eval = FALSE`) so the
vignette builds quickly --- flip them to `eval = TRUE` (and raise `n_sim`) to
reproduce them.

```{r library}
library(spaCraft)
```

## The pilot data

A pilot dataset is a plain `list` of samples. Each element is itself a list with
a `counts` matrix (genes \eqn{\times} spots), a `coords` data frame (with
columns `x`, `y`, and the annotated `domain`), a `group` label (control = `0`,
case = `1`), and a `sample_id`. This is exactly the format you would assemble
from a `SpatialExperiment`/`Seurat` object or from `spatialLIBD`. For your own
analysis you would load such a list and pass it straight to
`createspaCraftObject()`:

```{r real-data, eval = FALSE}
# For a real analysis: a list of samples, each with counts, coords (x, y,
# domain), group, and sample_id.
load("path/to/visium_human_brain_pilot_data_list.RData")  # provides `pilot_data_list`
spaCraft_obj <- createspaCraftObject(pilot_data_list)
```

Here we use the bundled, downsampled DLPFC pilot list and build the object from
it. This is the first and only place we touch a `mini_*` dataset --- everything
downstream grows out of `spaCraft_obj`.

```{r create-object}
data(visium_human_brain_pilot_data_list)

length(visium_human_brain_pilot_data_list)               # number of pilot samples
str(visium_human_brain_pilot_data_list[[1]], max.level = 1)

spaCraft_obj <- createspaCraftObject(visium_human_brain_pilot_data_list)
```

`createspaCraftObject()` validates the list and, if a group has only a single
replicate (\eqn{K = 1}), can optionally generate pseudo-replicates so that the
downstream multi-sample estimators remain well defined (see
`generate_pseudo_reps`). The result is the S4 object that every later step
reads from and writes to.

## Feature selection

`spaCraft` works with two complementary gene sets, identified in two passes.
`featureSelection()` performs a multi-context meta-analysis to find robust
**domain-informative markers** (`top_genes`), while `featureSelectionStable()`
identifies **spatially and group-wise stable genes** (`stable_genes`) that serve
as an empirical null and as the pool from which spike-in genes are later drawn.
Because the bundled data is downsampled, we deliberately **relax the cutoffs**
so that genes survive on a few hundred features; on full data you would use
stricter thresholds (e.g. `logfc_cutoff = 0.5`, `mean_in_cutoff = 1`).

```{r feature-selection}
spaCraft_obj <- featureSelection(
  object         = spaCraft_obj,
  logfc_cutoff   = 0.01,
  mean_in_cutoff = 0.01,
  max_num_gene   = 10,
  combine_method = "union",
  n_cores        = 1
)

spaCraft_obj <- featureSelectionStable(
  object              = spaCraft_obj,
  control_group_value = "0",
  case_group_value    = "1",
  check_group_diff    = TRUE,
  max_group_logfc_cut = 0.5,
  mean_global_cut     = 0.1,
  pct_in_cut          = 0.1,
  min_domain_prop     = 0.7,
  max_abs_logfc_cut   = 0.5,
  cv_cut              = 0.8,
  max_num_gene        = 50,
  use_low             = TRUE,
  n_cores             = 1,
  verbose             = TRUE
)

head(spaCraft_obj@params_expression$top_genes)
head(spaCraft_obj@params_expression$stable_genes)
```

**Why two passes?** The markers (`top_genes`) carry the spatial signal that lets
`pBANSKY` recover domains, whereas the stable genes (`stable_genes`) are
deliberately *un*informative --- flat across space and across groups --- so that
injecting a controlled effect into a few of them produces a clean, known
alternative against an honest null.

## Defining the simulation gene sets

`makeCustomGeneSets()` turns the two selected sets into the three gene sets the
simulator and the endpoints consume, for a chosen Target/Reference domain pair
(here white matter vs. layer 6):

- `G_svg` --- spatially variable genes (the markers plus stable genes) used to
  *label* spots;
- `G_null` --- the stable / empirical-null genes; and
- `G_spike` --- the top differentially expressed genes between Target and
  Reference, used for *signal injection*.

```{r gene-sets}
G_svg_base <- spaCraft_obj@params_expression$top_genes
G_stable   <- spaCraft_obj@params_expression$stable_genes

sets <- makeCustomGeneSets(
  object        = spaCraft_obj,
  G_svg_base    = G_svg_base,
  G_stable      = G_stable,
  target_domain = "WM",
  ref_domain    = "Layer6",
  sample_index  = 1L,
  n_de          = 5
)

lengths(sets[c("G_svg", "G_null", "G_spike")])
head(sets$TR_stats, n=3)
```

## Fitting the three-layer generative model

With gene sets in hand we fit the cohort-level model. The three layers are
independent calls and each writes into its own slot of `spaCraft_obj`.

**Expression (NNGP).** `estimateExpressionParams_v2()` estimates per-gene
spatial covariance and biological variance across samples using
nearest-neighbour Gaussian-process approximations (Saha & Datta, 2018),
accelerated with Rcpp.

```{r estimate-expression}
spaCraft_obj <- estimateExpressionParams_v2(
  object           = spaCraft_obj,
  target_domain    = "WM",
  reference_domain = "Layer6",
  genes_to_use     = sets$G_svg,
  n_neighbors      = 15,
  n_cores          = 1,
  verbose          = TRUE
)
```

**Geometry (FGKMM).** `estimateGeometryParams_v2()` pools each domain's spot
cloud across samples and fits a Fisher--Gaussian kernel mixture (Mukhopadhyay,
Li & Dunson, 2020), selecting the number of kernels by BIC over `M_candidates`.

```{r estimate-geometry}
spaCraft_obj <- estimateGeometryParams_v2(
  object       = spaCraft_obj,
  M_candidates = 3:6,
  iter_max     = 1000,
  tol          = 1e-2,
  n_cores      = 1,
  verbose      = TRUE
)
```

**Composition (Binomial--logit).** `estimateCompositionParams_v2()` fits the
baseline-anchored model for the Target-vs-Reference abundance; the pilot
baseline \eqn{\beta_1} becomes the null center used by the LOR TREAT test.

```{r estimate-composition}
spaCraft_obj <- estimateCompositionParams_v2(
  object           = spaCraft_obj,
  target_domain    = "WM",
  reference_domain = "Layer6",
  verbose          = TRUE
)
```

All three parameter slots are now populated. (This is exactly the state
captured by the bundled `mini_obj_fitted` object --- but here we built it
ourselves from the pilot list.)

```{r inspect-fitted}
names(spaCraft_obj@params_composition)   # baseline-anchored Binomial-logit
names(spaCraft_obj@params_expression)    # gene sets + NNGP parameters
names(spaCraft_obj@params_geometry)      # FGKMM geometry
```

We can visualize the pilot domains directly from the object. The S4 `plot()`
method draws spatial domains with a consistent palette across samples:

```{r plot-pilot, fig.alt = "Spatial domains of the pilot DLPFC samples"}
plot(spaCraft_obj, type = "pilot", n_per_group = 2)
```

## (Optional) Stability-driven hyperparameter tuning

The recover step has two spatial knobs: the conditional-texture strength of the
simulator (`lambda_cond`) and the BANKSY spatial-mixing weight of the clusterer
(`lambda_p`). `tuneSpaDesignLambdas()` selects them by stability under a single
\eqn{H_1} scenario, returning endpoint-specific optima for SaLFC and LOR. This
is a Monte-Carlo search, so it is shown here with small settings but **not run
during the build** (set `eval = TRUE` to run it).

```{r tuning, eval = TRUE}
tuning_H1_scenario <- list(
  DE_lfc = 0.3, target_prop_case = 0.7, kappa_geo = 1.0,
  kappa_m = 1.0, kappa_bio = 1.0, delta_rho = 1.0
)

tuning_obj <- tuneSpaDesignLambdas(
  object        = spaCraft_obj,
  target_domain = "WM", ref_domain = "Layer6",
  G_DE          = sets$G_spike,
  scenario_H1   = tuning_H1_scenario,
  lambda_p_grid = seq(0.3, 0.7, by = 0.1),   # small grid for speed
  lambda_c_grid = seq(0.3, 0.7, by = 0.1),
  B = 20, K_syn = 10, n_cores = 1, seed = 123
)

summary(tuning_obj)
plot(tuning_obj)
```

The values selected here (`tuning_obj$optimal_params$SaLFC$lambda_p`, etc.) can
be passed to the power functions below; if you skip tuning, just use fixed
values, as we do next.

## Simulating a synthetic cohort

One synthetic cohort makes the pipeline concrete. The scenario list controls the
effect being injected; here we turn expression DE *off* (`DE_lfc = 0`) and raise
the case-group target proportion to 0.7, i.e. a purely compositional effect.

```{r simulate}
scenario_params <- list(
  DE_lfc           = 0.3,   # expression log-fold-change for DE genes
  target_prop_case = NULL,   # case-group target-domain proportion
  kappa_geo = 1.0, kappa_m = 1.0, kappa_bio = 1.1,
  delta_rho = 1.0, lambda_cond = tuning_obj$optimal_params$SaLFC$lambda_cond
)

sim_results <- simulatespaCraft(
  object             = spaCraft_obj,
  n_sample_per_group = 5,
  scenario_settings  = scenario_params,
  target_domain      = "WM",
  genes_to_simulate  = sets$G_svg,
  de_genes           = sets$G_spike,
  spatial_mode       = "original",
  graph_k            = 30L,
  n_cores            = 1
)
```

## Recovering spatial domains and visualizing

`pBANSKY()` recovers domain labels on the synthetic samples using a
pilot-centroid-guided BANKSY embedding; the recovered labels are written to
`coords$hat_d`. We then re-shape the synthetic blobs onto the pilot geometry
with `rearrangeSyntheticToPilot()` and store the result back in the object.

```{r recover}
syn_d_hat <- pBANSKY(
  sim_data   = sim_results,
  pilot_data = spaCraft_obj@pilot_data,
  lambda = tuning_obj$optimal_params$SaLFC$lambda_p, k_neighbors = 30,
  scale_pilot_coords = FALSE, scale_sim_coords = FALSE,
  hard_guidance = FALSE, n_pcs = 20, do_hungarian = TRUE,
  use_G = TRUE, balance_mode = "energy_proportional"
)

# pack into the {counts, coords, meta} shape the downstream steps expect
synthetic_data <- setNames(
  lapply(syn_d_hat, function(item) {
    list(counts = item$counts, coords = item$coords,
         meta = data.frame(sample_id = item$sample_id, group = item$group))
  }),
  vapply(syn_d_hat, function(item) item$sample_id, character(1))
)

mapped_synthetic_data <- rearrangeSyntheticToPilot(
  synthetic_data   = synthetic_data,
  pilot_data       = spaCraft_obj@pilot_data,
  boundary_shift   = TRUE,          # render a composition shift as a moving boundary
  target_domain    = "WM",
  reference_domain = "Layer6"
)

syntheticData(spaCraft_obj) <- mapped_synthetic_data
```

**`lambda` and `boundary_shift`.** `lambda` is BANKSY's spatial-mixing weight
(0 = expression only, 1 = neighbourhood only); this is the `lambda_p` you would
tune. `boundary_shift = TRUE` is appropriate for the LOR (composition) endpoint:
it renders a larger case-group target proportion as the target band expanding
into the reference, leaving per-domain counts (and hence the LOR statistic)
untouched while only the spatial footprint changes.

Now that synthetic data are stored in the object, the same `plot()` method shows
the raw simulated domains and the pilot-mapped versions:

```{r plot-synthetic, fig.alt = "Synthetic domains before and after mapping onto pilot geometry"}
plot(spaCraft_obj, type = "synthetic_mapped",
     color_by = "domain", n_per_group = 3, cex = 1.2, alpha = 0.8)
```

## A single-design-point test

Before sweeping a grid, it is instructive to apply each endpoint **once** to the
single synthetic cohort we just recovered. This exposes the raw test output and
is exactly the computation that `evaluatePowerSaLFC()` and `evaluatePowerLOR()`
repeat internally at every \eqn{(K, \text{effect})} cell. We pass
`synthetic_data` --- the recovered cohort carrying cluster labels in
`coords$hat_d` --- and let both tests derive their TREAT thresholds directly
from the pilot (`lfc_threshold = "pilot"`, `tau = "pilot"`).

**SaLFC.** Run here on the null genes, so the p-values serve as a size check ---
recall that this cohort was simulated with `DE_lfc = 0` (a purely compositional
effect), so no expression signal was injected:

```{r onepoint-salfc}
res_SaLFC <- SaLFC(
  synthetic_data, spaCraft_obj,
  genes         = sets$G_null,
  target_domain = "WM", ref_domain = "Layer6",
  domain_col    = "hat_d",
  lfc_threshold = "pilot"
)

sets$G_null               # empirical-null genes
sets$G_spike              # injected-DE genes (relevant when DE_lfc > 0)
round(res_SaLFC$p_value, 5)
```

**LOR.** A single composition test for the WM-vs-Layer6 abundance; this is the
endpoint expected to register the injected `target_prop_case = 0.7` shift:

```{r onepoint-lor}
res_LOR <- LOR(
  synthetic_data, spaCraft_obj,
  target_domain = "WM", reference_domain = "Layer6",
  null_type = "treat", tau = "pilot"
)

res_LOR$p_value
```

Each rejection (`p_value <= alpha`) is a single Bernoulli outcome; the power
functions below simply average many such outcomes across the
\eqn{(K, \text{effect})} grid.

## Power analysis

We are now ready for the headline output: power as a function of the per-group
sample size \eqn{K}. Each grid point runs the full
simulate \eqn{\to} recover \eqn{\to} test pipeline `n_sim` times; the effect
\eqn{= 0} row is the empirical size/FDR. These sweeps are the expensive part of
the package, so they are shown here with small grids but **left unevaluated**;
set `eval = TRUE` and raise `n_sim` (e.g. 100--500) for a real study. We pass
fixed spatial weights below; in a full analysis you would substitute the tuned
values, e.g. `clustering_args = list(lambda = tuning_obj$optimal_params$SaLFC$lambda_p)`.

### SaLFC --- spatially adjusted differential expression

The swept effect is the expression log-fold-change `lfc_grid`. Pilot-based TREAT
thresholds are estimated once per effect size and shared across replicates.

```{r power-salfc-eval, eval = FALSE}
library(dplyr); library(ggplot2); library(ggrepel); library(scam)

# Note: This evaluation is computationally intensive.
power_results <- evaluatePowerSaLFC(
  object   = spaCraft_obj,
  K_grid   = seq(3, 9, by=1),
  lfc_grid = c(0.30, 0.40, 0.50),
  n_sim    = 10,
  seed     = 2026,
  G_svg    = sets$G_svg,
  G_null   = sets$G_null,
  G_spike  = sets$G_spike,
  scenario_base   = list(lambda_cond = tuning_obj$optimal_params$SaLFC$lambda_cond),
  clustering_args = list(lambda = tuning_obj$optimal_params$SaLFC$lambda_p)
)
```
```{r power-salfc-eval-curve, eval = TRUE}
library(dplyr); library(ggplot2); library(ggrepel); library(scam)

# Load pre-computed power results from the package to expedite vignette rendering
salfc_csv_path <- system.file("data", "Visium_human_brain_power_results_SaLFC.csv", package = "spaCraft")
power_results <- read.csv(salfc_csv_path)

print(head(power_results))                             # compact power matrix
plotPowerCurveSaLFC(power_results, alpha = 0.05)       # SCAM-smoothed curves
```

### LOR --- compositional shift

Here the swept effect is the case-group target proportion (`effect_type =
"target_prop_case"`); expression DE is switched off internally so the signal is
purely compositional. LOR returns a single test per replicate, so power is
simply the rejection rate.

```{r power-lor, eval = FALSE}
# Note: This evaluation is computationally intensive.
lor_results <- evaluatePowerLOR(
  object      = spaCraft_obj,
  K_grid      = seq(3, 9, by=1),
  delta_grid  = c(0.60, 0.70, 0.80),
  effect_type = "target_prop_case",
  n_sim       = 10,
  seed        = 2026,
  genes       = sets$G_svg,
  scenario_base   = list(lambda_cond = tuning_obj$optimal_params$LOR$lambda_cond),
  clustering_args = list(lambda = tuning_obj$optimal_params$LOR$lambda_p)
)
```

```{r power-lor-curve, eval=TRUE}
# Load pre-computed LOR results from the package
lor_csv_path <- system.file("data", "Visium_human_brain_power_results_LOR.csv", package = "spaCraft")
lor_results <- read.csv(lor_csv_path)

print(head(lor_results))
plotPowerCurveLOR(lor_results, alpha = 0.05)
```

**Reading the curves.** Each line is one effect size; the dashed reference line
is the nominal level \eqn{\alpha}. The `summary()` reports, for each (non-null)
effect, the smallest \eqn{K} reaching the target power --- the per-group sample
size you would recommend --- together with a size/FDR diagnostic at the null.

## Session info

```{r session-info}
sessionInfo()
```
